---
title  : 'Programming assignment 4'
output: github_document
editor_options: 
  chunk_output_type: console
---

**Author**: Alejandro Jaume-Losa
**Date**: Last update: `r Sys.time()`


# Overview

## What did I do?

In order to complete this last programming assignment, I first forked the repo and cloned it to my computer using Github Desktop. Then, I took a look at its structure and the README files in each folder. Once I had a "clear" idea of what I had to (+/-), I proceed to run the first Praat script. This script was intended to normalize the peak intensity of all of the files in the 'wavs' folders. This script was not broken and I did not have to add anyting else either, so I just ran it. Once I was done doing that, I proceeded to run the second script, which was intended to read in every .wav file from the 'wavs' folders and create a corresponding TextGrid with the same name. This script had some missing information, so I had to add the participant ID (i.e., bi02) and also replace “XXXX” with “45” (i.e., the total number of recordings per participant). Once that was done, I proceeded to segment all the file using the TextGrids created by the second Praat script (this took a while). Finally, once I was done doing that, I proceeded to run the third and last Praat script, which was intended to extract the VOT, F1, and F2 values from each file and create a .csv file for each participant. Once I had all the .csv files, I opened R and loaded the necessary packages. Then, I loaded the six .csv files and combined them. After that, I tidyed the data, I did some descriptive stats, and created some plots.

## Hypothesis

For this programming assignment, we are working with data from 3 bilingual speakers (Spanish-English), and 3 L2 learners (native speakers of English). Based on this, we can assume that their production of the target words will be different. For instance, in terms of VOT, Spanish and English are different: Spanish has short and lead lag, while English has short, long, and lead lag). This can be summarized by saying that the VOT for Spanish stops is shorter compared to the longer VOT that English stops have. Therefore, my hypothesis is that the L2 learners will have longer VOTs, due to influence of English.

# Prep

## Libraries

```{r}
#| label: load-libs
#| echo: false 
#| warning: false 
#| message: false
# Load libraries here
library(dplyr)
library(ggplot2)
library(readr)
library(here)
library(tidyverse)
library(tidyr)
library(knitr)
library(stringr)
library(praatpicture)
```

## Load data

```{r}
#| label: load-data
# You need to get all the files in the 'data' directory and combine them
# Check previous examples we did in class

read_csv(file = "../data/bi01.csv")
read_csv(file = "../data/bi02.csv")
read_csv(file = "../data/bi03.csv")
read_csv(file = "../data/ne01.csv")
read_csv(file = "../data/ne02.csv")
read_csv(file = "../data/ne03.csv")

bi01 <- read_csv(here("data", "bi01.csv"))
bi02 <- read_csv(here("data", "bi02.csv"))
bi03 <- read_csv(here("data", "bi03.csv"))
ne01 <- read_csv(here("data", "ne01.csv"))
ne02 <- read_csv(here("data", "ne02.csv"))
ne03 <- read_csv(here("data", "ne03.csv"))

data <- rbind(bi01,bi02,bi03, ne01,ne02, ne03)

bi <- rbind(bi01,bi02,bi03)
ne <- rbind(ne01,ne02, ne03)

```

## Tidy data

```{r}
#| label: tidy-data
# Convert from long to wide or wide to long format as necessary (check 
# examples from class)
# Create any other relevant variables here 

# Tidying data

data_tidy <- data %>%
  separate(fileID, into = c("id","stim"), sep = 4) %>%
  mutate(word = str_remove(stim, "[_]")) %>%
  mutate(item = str_remove(word, "[1-2]")) %>%
  mutate(group = case_when(
    startsWith(id, "bi") ~ "bi",
    startsWith(id, "ne") ~ "ne"
  ))

bi_tidy <- bi %>%
  separate(fileID, into = c("id","stim"), sep = 4)%>%
  mutate(word = str_remove(stim,"[_]"))%>%
  mutate(item = str_remove(word, "[1-2]"))

ne_tidy <- ne %>%
  separate(fileID, into = c("id","stim"), sep = 4)%>%
  mutate(word = str_remove(stim,"[_]"))%>%
  mutate(item = str_remove(word, "[1-2]"))

# Getting final data sets

data_final <- select(data_tidy, id, group, item, vot, f1, f2)

bi_final <- select(bi_tidy, id, item, vot, f1, f2)

ne_final <- select(ne_tidy, id, item, vot, f1, f2)

```

# Analysis

## Descriptives

```{r}
#| label: descriptive-stats
# Give some descriptive summaries of your data 
# Display your descriptives in a table (try knitr::kable())

means <- data_final %>%
  group_by(id) %>%
  summarise(vot_mean = mean(vot), vot_sd = sd(vot), f1_mean = mean(f1), f1_sd = sd(f1), f2_mean = mean(f2), f2_sd = sd(f2))

kable(means)

means_group <- data_final %>%
  group_by(group) %>%
  summarise(vot_mean = mean(vot), vot_sd = sd(vot), f1_mean = mean(f1), f1_sd = sd(f1), f2_mean = mean(f2), f2_sd = sd(f2))

kable(means_group)

bi_means <- bi_final %>%
  group_by(id) %>%
  summarise(vot_mean = mean(vot), vot_sd = sd(vot), f1_mean = mean(f1), f1_sd = sd(f1), f2_mean = mean(f2), f2_sd = sd(f2))

kable(bi_means)

ne_means <- ne_final %>%
  group_by(id) %>%
  summarise(vot_mean = mean(vot), vot_sd = sd(vot), f1_mean = mean(f1), f1_sd = sd(f1), f2_mean = mean(f2), f2_sd = sd(f2))

kable(ne_means)

means_vot_group <- data_final %>%
  group_by(group) %>%
  summarise(vot_mean = mean(vot), vot_sd = sd(vot))

kable(means_vot_group)

```

## Visualization

```{r}
#| label: plots 
#| fig.retina: 2
# Include some plots here

# VOT as a function of item

data_final %>%
  ggplot(aes(x = vot, y = item, color = id)) +
  geom_point() +
  facet_grid(~ group)

# F1 as a function of item

data_final %>%
  ggplot(aes(x = f1, y = item, color = id)) +
  geom_point() +
  facet_grid(~ group)

# F2 as a function of item

data_final %>%
  ggplot(aes(x = f2, y = item, color = id)) +
  geom_point() +
  facet_grid(~ group)

# VOT as a function of participant

data_final %>%
  ggplot(aes(x = id, y = vot, fill = group)) +
  geom_boxplot()

# F1 as a function of participant

data_final %>%
  ggplot(aes(x = id, y = f1, fill = group)) +
  geom_boxplot()

# F2 as a function of participant

data_final %>%
  ggplot(aes(x = id, y = f2, fill = group)) +
  geom_boxplot()

# VOT as a function of stop type (and vowel) and group

plot_va = filter(data_final, item == "kaka" | item == "paka" | item == "taka") 
ggplot(plot_va, aes(x = item, y = vot, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_ve = filter(data_final, item == "keke" | item == "peke" | item == "teke") 
ggplot(plot_ve, aes(x = item, y = vot, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_vi = filter(data_final, item == "kiki" | item == "piki" | item == "tiki") 
ggplot(plot_vi, aes(x = item, y = vot, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_vo = filter(data_final, item == "koko" | item == "poko" | item == "toko") 
ggplot(plot_vo, aes(x = item, y = vot, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_vu = filter(data_final, item == "kuku" | item == "puku" | item == "tuku") 
ggplot(plot_vu, aes(x = item, y = vot, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)


# F1 as a function of stop type (and vowel) and group

plot_f1a = filter(data_final, item == "kaka" | item == "paka" | item == "taka") 
ggplot(plot_f1a, aes(x = item, y = f1, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f1e = filter(data_final, item == "keke" | item == "peke" | item == "teke") 
ggplot(plot_f1e, aes(x = item, y = f1, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f1i = filter(data_final, item == "kiki" | item == "piki" | item == "tiki") 
ggplot(plot_f1i, aes(x = item, y = f1, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f1o = filter(data_final, item == "koko" | item == "poko" | item == "toko") 
ggplot(plot_f1o, aes(x = item, y = f1, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f1u = filter(data_final, item == "kuku" | item == "puku" | item == "tuku") 
ggplot(plot_f1u, aes(x = item, y = f1, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

# F2 as a function of stop type (and vowel) and group

plot_f2a = filter(data_final, item == "kaka" | item == "paka" | item == "taka") 
ggplot(plot_f2a, aes(x = item, y = f2, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f2e = filter(data_final, item == "keke" | item == "peke" | item == "teke") 
ggplot(plot_f2e, aes(x = item, y = f2, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f2i = filter(data_final, item == "kiki" | item == "piki" | item == "tiki") 
ggplot(plot_f2i, aes(x = item, y = f2, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f2o = filter(data_final, item == "koko" | item == "poko" | item == "toko") 
ggplot(plot_f2o, aes(x = item, y = f2, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)

plot_f2u = filter(data_final, item == "kuku" | item == "puku" | item == "tuku") 
ggplot(plot_f2u, aes(x = item, y = f2, fill = group)) + 
  geom_boxplot() +
  facet_wrap(~ group)


```

# Praat pictures

```{r}

# Wave form and spectogram of a bilingual participant

praatpicture::praatpicture("bi01_kaka.wav")

# Wave form and spectogram of a native speaker (L2 learner) participant

praatpicture::praatpicture("ne01_kaka.wav")


```


## Hypothesis test

```{r}
#| label: stats
# Conduct a simple statistical analysis here (optional)

```


# Conclusion

The initial hypothesis was that the L2 learners will have longer VOTs due to the influence of their L1. After analyzing the data, specifically the VOT averages for bilingual participants compared to the VOT averages for L2 learners, we can see that each individual L2 learner had longer VOT than the bilingual participants.

```{r}
kable(bi_means)
kable(ne_means)
```

Additionally, taken as a whole, the L2 learner group also had longer VOT than the bilingual group.

```{r}
kable(means_vot_group)
```

# Reflections

This was a VERY long assignment. In my opinion, the most complicated part was understanding what had to be done every time. I felt like the instruction were not as detailed compared to the three previous programming assignments and I never really knew if I was doing what was supposed to be done. Also, segmenting all the files took a very long time. The R part was relatively easy. However, I believe this assignment was very complete. I was able to apply all the skills I learned in this class.

</br></br>
